---
title: "Using mtrostR for analysis of fatal injuries suffered in motor vehicle traffic crashes"
author: "Struya"
date: "`r Sys.Date()`"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Using mtrostR for analysis of fatal injuries suffered in motor vehicle traffic crashes}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>"
)
```

```{r, include = FALSE}
library(mtrostR)
```

Database on fatal injuries suffered in motor vehicle traffic crashes is  produced by [US National Highway Traffic Safety Administration](https://www.nhtsa.gov/). Data is provided on a yearly frequency. 
Using functions from the `mtrostR` package, one can import data from files disseminated by Administration, summarise it and plot fatal injuries by State on a map.

## Installing `mtrostR`

Package `mtrostR` is only available on GitHub. It can be installed like:

```{r, echo = TRUE, eval = FALSE}
# install.packages("devtools")
devtools::install_github("Struya/mtrostR")
```

## Getting started

### Construct filename

Using this package function `make_filename` one can construct the filename string to be used for importing the dataset into R session. The user must only provide a number for a year (i.e. 2013) to be used by the function. Please note that in the package only dataset from 2013 to 2015 are included. 

```{r}
# choose a year
my_year <- 2013

# construct filename to be used for importing dataset into R
my_filename <- make_filename(year = my_year)
```

### Import dataset

Importing fatalities dataset is simple. Using `fars_read` with filename string one can import a dataset for a chosen year.

```{r}
# import datast into R using filename string
my_df <- fars_read(my_filename)

my_df
```

There is also a function `fars_read_years` that imports datasets for all years at once but returns only two columns, month and year. The user only has to provide a vector of years. 

```{r}
# vector of years 
my_years <- 2013:2015

# import all datasets into R. 
my_list <- fars_read_years(my_years)

my_list
```

## Summarizing datasets

Using `fars_read_years` to import only columns months and years might be useful, however there is a function that calls it only to import datasets in order to summarise it by year and month. It is named `fars_summarise_years`.

```{r}
# vector of years
my_years <- 2013:2015

# count number of fatalities by month and year
count_by_year_and_month_df <- fars_summarize_years(my_years)
count_by_year_and_month_df
```

## Figures

Plotting fatalities on a map is a nice way of getting a feel for spatial distribution of fatalities accross United States. By providing the State ID number and a year, the function `fars_map_state` will plot spatial distribution on a map of chosen State. 

```{r, fig.show= 'hold'}
# choice of State
my_state_id <- 1
my_year <- 2015

my_state_id_to_compare <- 10 

# plot spatial distribution for a year
fars_map_state(my_state_id, my_year)
fars_map_state(my_state_id_to_compare, my_year)
```
